---
title: "SLEEP PATTERNS BY DIET"
author: "GRACY CHANGIRALA"
date: "`r Sys.Date()`"
output:
  word_document: 
    toc: true
    fig_caption: true
    keep_md: true
    fig_width: 7
    fig_height: 5.5
  pdf_document: default
  html_document: 
    toc: true
    fig_caption: true
    number_sections: true
---
# Loading the libraries and Viewing Data
```{r}
library(tidyverse)
data()
view(msleep)
```
*This code loads the 'tidyverse', package, which is a collection of packages in R for Data manipulation and visualisation. The 'data()' function lists all available datasets in R, and 'view(msleep)' opens the 'msleep' dataset in a new tab for inspections.*


# Descriptive Statistics on 'msleep' Data
```{r}
# Describe the spread, centrality and variance of the data
min(msleep$awake)
max(msleep$awake)
range(msleep$awake)
IQR(msleep$awake)
mean(msleep$awake)
median(msleep$awake)
var(msleep$awake)
```
**These statistics give us the insights into the distribution and variability of the 'awake' variable.**

*This block of code calculates basic descriptive statistics for the 'awake' variable in the 'msleep' dataset:*

**- 'min()', 'max()', 'range()'**: *Calculates the minumum, maximum and ranges of the dataset.*

**- 'IQR()'**: *Interquartile range, showing the spread of the middle 50% of the data.*

**-'mean()', 'median()'"**: *Central tendency measures.*

**- 'var()'**: *Variance, indicating the spread of the data.*



# Summary Statistics for the Entire Dataset
```{r}
#Summarise selected variables
summary(msleep)
summary(msleep$sleep_total)

msleep %>% 
  select(sleep_total, brainwt) %>%
  summary
```
**Summary statistics gives and overview of each variable, helping identify trends, outliners, and missing values.**


**- 'summary(msleep)'**: *Provides summary statistics for all variable in the 'msleep' dataset.*

**- 'summary(msleep$sleep_total)'**: *Specifically summarizes the 'sleep_total' variable.*

**- 'select(sleep_total, brainwt) %>% summary'**: *Selects the 'sleep_total' and 'brainwt' variables and then summarizes them.*

# Creating a summary Table by Grouping
```{r}
# Create a summary table
msleep %>%
  drop_na(vore) %>%
  group_by(vore) %>%
  summarise(
    Lower = min(sleep_total),
    Average = mean(sleep_total),
    Upper = max(sleep_total),
    Difference = max(sleep_total) - min(sleep_total)
  ) %>%
  arrange(Average)
```
**Grouping and summarising data this way helps compare different categories and understand how they differ in terms of sleep patterns.**

*This code creates a summary table for 'sleep_total' based on the 'vore'(dietary category) variable:*

**- 'drop_na(vore)'**: *Remove rows where 'vore' is 'NA'.*

**- 'group_by(vore)'**: *'Groups data by the 'vore' category.*

**- 'summarise()'**: *Calculates the minimum, average, maximum, and range ('Difference') of 'sleep_total' within each 'vore' category.*

**- arrange(Average)'**" *Sorts the results by the average 'sleep_total'.*

# Contingency Tables
```{r}
# Creating contingency tables
library(MASS)
attach(Cars93)

glimpse(Cars93)

table(Origin)
table(AirBags, Origin)
addmargins(table(AirBags, Origin))

table(AirBags, Origin)
prop.table(table(AirBags, Origin))*100
round(prop.table(table(AirBags, Origin), 2)*100)
```

**Contingency tables are useful for exploring relationships between categorical variable.**

**- attach(Cars93)**: *Makes the variables in the Cars93 dataset accessible directly.*

**-  glimpse(Cars93)**: *Provides a quick look at the Cars93 dataset structure.*

**-  table(Origin)**: *Creates a frequency table for the Origin variable.*

**-  table(AirBags, Origin)**: *Creates a contingency table showing the relationship between AirBags and Origin.*

**-  addmargins(table(AirBags, Origin))**: *Adds row and column sums (margins) to the contingency table.*

**-  prop.table()**: *Converts the contingency table into proportions, showing the percentage of each combination.*


# Pivoting Data for Summary
```{r}
Cars93 %>%
  group_by(Origin, AirBags) %>%
  summarise(number = n()) %>%
  pivot_wider(names_from = Origin,
              values_from = number)
```
**Turning helps in reshaping information, making it more straightforward to picture and analyze bunches next to each other.**

**This code block groups the Cars93 dataset by Origin and AirBags, counts the number of occurrences (n()), and then uses pivot_wider to reshape the data:**

**- group_by(Origin, AirBags)**: *Groups the data by Origin and AirBags.*

**- summarise(number = n())**: *Counts the number of cars in each group.*

**- pivot_wider()**: *Converts the long-format data into a wider format where the Origin categories become separate columns.*


# Conclusion

*In the wake of running the above code blocks, you ought to have a complete synopsis and depiction of the msleep and Cars93 datasets. The examination ought to give experiences into the circulation, focal inclination, fluctuation, and connections inside the information.*

*This sort of examination is essential in information science, assisting you with understanding your information prior to applying more complicated models or representations.*



